Exploring the Demographic Aspects of COVID-19 Cases in Toronto: Are the elderly males more susceptible to COVID-19?

### 2024-01-26 outputs/paper/paper.qmd
Update paper, simulate data, download data, data cleaning, README, llm usage, and add test data based on the peer review suggestions
Fix typos, add pre-requisites, additional comments to the code, and add details to the findings

Add tests for the cleaned data

The dataset 'COVID-19 Cases in Toronto'  [@covidCases] was utilized and obtained from 'opendatatoronto' library [@openDataToronto]. It comprises 412,122 records and 15 variables, including assigned ID, age group, neighborhood name, episode date, client gender, and more within the COVID-19 cases in Toronto. The data cover the period from January 22, 2020, to January 4, 2024, and are structured in CSV file format.

The primary variables selected for analysis were chosen based on their relevance to my research question, relating to age and gender. If one categorical value was missing, then the entire row was removed Therefore, the cleaned COVID-19 cases dataset now includes three variables: case ID, gender, and age group (see @tbl-idcasetable). The total number of rows has been reduced from 412,122 to 411,700.

@@

head(cleaned_case_data) |>
  kable(
    col.names = c("Case ID", "Age group", "Gender")
  )
```

@@

# Reference: https://tellingstorieswithdata.com/02-drinking_from_a_fire_hose.html#explore 
analyzed_case_data |> 
  ggplot(aes(x = gender)) +
  geom_bar() +
  geom_text(
    stat = 'count',
    aes(label = scales::percent(..count.. / sum(..count..)),
        y = ..count..),
    vjust = -0.5,
    position = position_stack(vjust = 0.5)
  ) +
  theme_minimal() +
  labs(x = "Gender",
       y = "Number of covid-19 cases")
```

## COVID-19 Cases by Age Groups

@@

```{r}
#| label: fig-casebyagegroup
#| fig-cap: Number of covid-19 cases by age groups
#| echo: false
#| warning: false
#| message: false

# Reference: https://tellingstorieswithdata.com/02-drinking_from_a_fire_hose.html#explore 
analyzed_case_data |> 
  ggplot(aes(x = age_group)) +
  geom_bar() +
  geom_text(
    stat = 'count',
    aes(label = scales::percent(..count.. / sum(..count..)),
        y = ..count..),
    vjust = -0.5,
    position = position_stack(vjust = 0.5)
  ) +
  theme_minimal() +
  labs(x = "Age groups",
       y = "Number of covid-19 cases")
```

@@

# Reference: https://rstudio-pubs-static.s3.amazonaws.com/215516_c2cee6b2885d419e9248e7eb7c87f887.html
analyzed_case_data |> 
  ggplot(aes(x = age_group, fill = gender)) +
  geom_bar(position = "dodge") +
  scale_x_discrete("Age groups") +
  scale_fill_discrete(name = "Gender",
                      breaks = c("Male", "Female", "Other")) +
  theme_minimal() +
  labs(y = "Number of covid-19 cases")
```

@@

## Interpretation of Findings
The examination of COVID-19 cases by gender reveals that females represent 53.4% of the cases in Toronto and males represent 45.7% (see @fig-casebygender), which does not show significant differences between female and male cases. The age-stratified analysis of COVID-19 cases shows that the 20-29 age group is most affected by COVID-19, accounting for 18.83% of the overall cases (see @fig-casebyagegroup). Conversely, older age groups (40+) have lower case numbers. Moreover, the intersectional analysis of gender and age groups shows that most age groups have a higher number of female cases, with exceptions in the 0-19 and 70-79 age groups where male cases are higher than female cases (see @fig-distribution). These findings do not support the hypothetical question: "Are elderly males more susceptible to COVID-19?" Additional demographic variables, such as socioeconomic status and cultural factors, may influence COVID-19 patterns in diverse populations.

@@

Future research could address these limitations by exploring other demographic variables and considering a wider geographical scope. This exploration would provide a more comprehensive understanding of the factors that influence the patterns of COVID-19 cases across diverse populations. Future research endeavors have the potential to refine public health strategies, tailor interventions to specific population needs, and contribute substantially to the global understanding of the multifaceted nature of the COVID-19 pandemic.

>>>>

The dataset 'COVID-19 Cases in Toronto'  [@covidCases] was utilized and obtained from 'opendatatoronto' library [@openDataToronto]. It comprises 412,122 records and 15 variables, including assigned ID, age group, neighborhood name, episode date, client gender, and more. The data cover the period from January 22, 2020, to January 4, 2024, and are structured in CSV file format.

The primary variables selected for analysis were chosen based on their relevance to my research question, relating to age and gender. If one categorical value was missing, then the entire row was removed. Therefore, the cleaned COVID-19 cases dataset now includes three variables: case ID, gender, and age group (see @tbl-idcasetable). The total number of rows has been reduced from 412,122 to 411,700.

@@

# Display the first six rows of the "cleaned_case_data" dataset
head(cleaned_case_data) |>
  kable(
    # Give each column a name
    col.names = c("Case ID", "Age group", "Gender")
  )
```

@@

# Create a bar chart to visualize the distribution of COVID-19 cases by gender based on the analyzed_case_data dataset
# Reference: https://tellingstorieswithdata.com/02-drinking_from_a_fire_hose.html#explore 
analyzed_case_data |> 
  ggplot(aes(x = gender)) +
  # Change bar color to green
  geom_bar(fill = "#D4E7C5") +
  geom_text(
    # Use the 'count' statistic to calculate counts for each category
    stat = 'count',
    # Use percentage labels based on count and set y position of text labels
    aes(label = scales::percent(..count.. / sum(..count..)),
        y = ..count..),
    # Adjust vertical position of text labels
    vjust = -0.5,
    # Stack text labels on top of bars
    position = position_stack(vjust = 0.5)
  ) +
  # Apply a minimal theme to the chart
  theme_minimal() +
  # Label x-axis as "Gender" and y-axis as "Number of COVID-19 cases"
  labs(x = "Gender",
       y = "Number of COVID-19 cases")
```

## COVID-19 Cases by Age Group

@@

```{r}
#| label: fig-casebyagegroup
#| fig-cap: Number of covid-19 cases by age group
#| echo: false
#| warning: false
#| message: false

# Create a bar chart to visualize the distribution of COVID-19 cases by age group based on the analyzed_case_data dataset
# Reference: https://tellingstorieswithdata.com/02-drinking_from_a_fire_hose.html#explore 
analyzed_case_data |> 
  ggplot(aes(x = age_group)) +
  # Change bar color to green
  geom_bar(fill = "#D4E7C5") +
  geom_text(
    # Use the 'count' statistic to calculate counts for each category
    stat = 'count',
    # Use percentage labels based on count and set y position of text labels
    aes(label = scales::percent(..count.. / sum(..count..)),
        y = ..count..),
    # Adjust vertical position of text labels
    vjust = -0.5,
    # Stack text labels on top of bars
    position = position_stack(vjust = 0.5),
    # Change text size to 3
    size = 3
  ) +
  # Apply a minimal theme to the chart
  theme_minimal() +
  # Label x-axis as "Age group" and y-axis as "Number of COVID-19 cases"
  labs(x = "Age group",
       y = "Number of COVID-19 cases")
```

@@

# Reference: https://rstudio-pubs-static.s3.amazonaws.com/215516_c2cee6b2885d419e9248e7eb7c87f887.html
analyzed_case_data |> 
  # Create a ggplot object with age_group on the x-axis and gender as fill color
  ggplot(aes(x = age_group, fill = gender)) +
  # Add a grouped bar layer to create the grouped bar chart
  geom_bar(position = "dodge") +
  # Scale the x-axis labels to display "Age group"
  scale_x_discrete("Age group") +
  # Scale the fill color to display "Gender" and specify the break points: "Male", "Female", "Other"
  scale_fill_discrete(name = "Gender",
                      breaks = c("Male", "Female", "Other")) +
  # Apply a minimal theme to the chart
  theme_minimal() +
  # Label y-axis as "Number of COVID-19 cases"
  labs(y = "Number of COVID-19 cases")
```

@@

## Interpretation of Findings
The examination of COVID-19 cases by gender reveals that females represent 53.4% of the cases in Toronto and males represent 45.7% (see @fig-casebygender), which does not show significant differences between female and male cases. This indicates that both genders are equally affected by COVID-19 in Toronto. Although the overall proportions may appear similar, there may be differences in the severity of the cases with each gender that need further investigation.

The age-stratified analysis of COVID-19 cases shows that the 20-29 age group is most affected by COVID-19, accounting for 18.83% of the overall cases (see @fig-casebyagegroup). Conversely, older age groups, particularly those aged 40 and above, have lower case numbers. This suggests a potential age-related pattern in COVID-19 susceptibility or exposure and highlights the importance of targeted age-specific interventions and public health measures to effectively mitigate the virus's spread.

Moreover, the intersectional analysis of gender and age groups shows that most age groups have a higher number of female cases, with exceptions in the 0-19 and 70-79 age groups where male cases are higher than female cases (see @fig-distribution). This highlights the complexity of COVID-19 epidemiology. Further research is needed to explore the underlying factors that cause this pattern.

These findings do not support the hypothetical question: "Are elderly males more susceptible to COVID-19?" Instead, the data suggests that COVID-19 patterns are influenced by factors beyond age and gender demographics. Additional demographic variables, such as socioeconomic status and cultural factors, may influence COVID-19 patterns in diverse populations.

@@

Future research could address these limitations by exploring other demographic variables and considering a wider geographical scope. This exploration would provide a more detailed understanding of the factors that influence the patterns of COVID-19 cases across diverse populations. This could have the potential to refine public health strategies, inform targeted public health interventions, and contribute substantially to the global understanding of the COVID-19 pandemic.
